---
title: "Non-parametric Tests"
format: pdf
---

```{r}
#| message: false

library(tidyverse)
library(BSDA) #for the sign test function
```

## Checking Normality

Example of marine biomass ratio taken from 32 marine organisms. Biomass ratio is used to test the protection effect of biological reserves used to conserve marine species.It is the total mass of all marine plants and animals per unit area of reserve divided by the same quantity in the unprotected control. If the biomass ratio equals one, than the reserve has no protection had no effect.

```{r}
#| fig.height: 4

# define data
bio_ratio = c(1.34, 1.96, 2.49, 1.27, 1.19, 1.15, 1.29, 1.05, 1.10, 1.21, 1.31,
              1.26, 1.38, 1.49, 1.84, 1.84, 3.06, 2.65, 4.25, 3.35, 2.55, 1.72,
              1.52, 1.49, 1.67, 1.78, 1.71, 1.88, 0.83, 1.16, 1.31, 1.40)

# Histogram - notice the severe right skew
hist(bio_ratio, xlab="Biomass Ratio", freq=T, col=2)

# Create a quantile-quantile plot (QQplot)
qqnorm(bio_ratio, col=2, pch=19, cex=1.5)

# Add a straight line which passes through the first and third quartiles.
qqline(bio_ratio, col = 1,lwd=2,lty=2)


# Perform Shapiro-Wilk test
shapiro.test(bio_ratio)
# reject the null, again evidence towards non-normality
```


## Non-parametric Sign Test

This test uses the signs of the differences, but not the magnitudes.          

$H_0:$ the median of the diff distribution is zero 

Test the effect of linoleic acid supplementation on systolic blood pressure. A group of 17 adults with a diet high in linoleic acid were assessed at baseline and 4 weeks later.

*Note: You will need to install the BSDA pacakge in order to use the function `SIGN.test`.  `install.packages("BSDA")`*

```{r}
# define data
base_SBP = c(119.67, 100, 123.56, 109.89, 96.22, 133.33, 115.78, 126.39,
             122.78, 117.44, 111.33, 117.33, 120.67, 131.67, 92.39, 134.44,
             108.67)
post_SBP = c(117.33, 98.78, 123.83, 107.67, 95.67, 128.89, 113.22, 121.56,
             126.33, 110.39, 107, 108.44, 117, 126.89, 93.06, 126.67, 108.67)

# perform test
SIGN.test(base_SBP, post_SBP, md=0)
```

## Non-parametric Wilcoxon-Signed Rank Test

```{r}
#| fig.height: 4

# Calculate the differences
diff_SBP = base_SBP-post_SBP

hist(diff_SBP)

# perform test
shapiro.test(diff_SBP)

qqnorm(diff_SBP, col=2, pch=19, cex=1.5)
qqline(diff_SBP, col = 1,lwd=2,lty=2)

# perform test
wilcox.test(diff_SBP)
# equivalently...
wilcox.test(base_SBP, post_SBP, paired=T) 
```
## Non-parametric Wilcoxon-Rank Sum Test: Two Independent Groups

Compare the length of hospital stay for patients with the same diagnosis admitted at two different hospitals.


```{r}
#| fig.height: 7


# define data
hosp1 = c(21,10,32,60,8,44,29,5,13,26,33)
hosp2 = c(86,27,10,68,87,76,125,60,35,73,96,44,238)

# check normality
par(mfrow=c(2,2))
hist(hosp1, xlab="Length of stay (days)", freq=T, main="Hospital1")

hist(hosp2, xlab="Length of stay (days)", freq=T, main="Hospital2")

qqnorm(hosp1, col=2, pch=19, cex=1.5)
qqline(hosp1, col = 1,lwd=2,lty=2)

qqnorm(hosp2, col=2, pch=19, cex=1.5)
qqline(hosp2, col = 1,lwd=2,lty=2)

# perform test
wilcox.test(hosp1, hosp2, mu=0)
```

Notice that the test statistic T1 is different from what we got in the slides (83.5)
Test statistic: W = 17.5, p-value = 0.001925

Why is that?
For Wilcoxon Rank Sum test only, R is calculating the statistic without n1(n1+1)/2 term. 

There is a note in the function documentation: 
"The literature is not unanimous about the definitions of the Wilcoxon rank sum and Mann-Whitney tests. The two most common definitions correspond to the sum of the ranks of the first sample with the minimum value subtracted or not: R subtracts and S-PLUS does not, giving a value which is larger by m(m+1)/2 for a first sample of size m."

If we want to get the same value of T1 per our formula in the slides, we need to add the n1(n1+1)/2 term. 

```{r}
res = wilcox.test(hosp1, hosp2, mu=0)

res$statistic = res$statistic + 11*(11+1)/2

res  
```


