---
title: "fisher_mcnemar"
format: pdf
---

```{r}
#| message: false
library(tidyverse)
```

# Fisher's Exact Test

## Example 1

Tea-time experiment.

```{r}
# define data
tea_exp = matrix(c(3,1,1,3),nrow=2)

# one-tailed test
fisher.test(tea_exp, alternative = "greater")

# two-sided test
fisher.test(tea_exp)

# use chi-squared test
chisq.test(tea_exp)

```

Notice that the two-sided p-value from Fisher is greater than the one generated by chi-square. This supports the conclusion that Fisher Exat Test is more conservative (harder to reject)

## Example 2

```{r}
# define data
practice_data = matrix(c(1,8,9,3), nrow=2,
                      dimnames=list(c("diet","non-diet"), c("men","women")))

# perform chi-squared
chisq.test(practice_data)

# fishers test
fisher.test(practice_data)

```


# Mcnemar Test for Binomial Matched-Pair Data

Two procedures are tested on the same 75 subjects in order to identify the absence/presence of the disease.

```{r}
# define data
procedure_data = matrix(c(41,8,14,12), nrow=2, byrow=T,
                       dimnames=list(c("positive","negative"),c("positive","negative")))

# perform test
mcnemar.test(procedure_data)

# What if you performed a chi-squared test instead?
chisq.test(procedure_data)

```

